---
title: "Pml project (datasci)"
author: "Owen Larson"
date: "1/25/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Overview
This is the final report for Coursera’s Practical Machine Learning course, as part of the Data Science Specialization track offered by John Hopkins.

In this project, we will use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants to predict the manner in which they did the exercise. This is the “classe” variable in the training set. We train 4 models: Decision Tree, Random Forest, Gradient Boosted Trees, Support Vector Machine using k-folds cross validation on the training set. We then predict using a validation set randomly selected from the training csv data to obtain the accuracy and out of sample error rate. Based on those numbers, we decide on the best model, and use it to predict 20 cases using the test csv set.

## Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: (http://groupware.les.inf.puc-rio.br/har) (see the section on the Weight Lifting Exercise Dataset).

Training data:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

Test data:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

downloaded: 1/25/22

The data for this project come from this source: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.

## Load libraries and data
```{r code}
library(lattice)
library(ggplot2)
library(kernlab)
library(rattle)
library(corrplot)
library(caret)
library(randomForest) ;library(gbm)
set.seed(1234)

traincsv <- read.csv("pml-training.csv")
testcsv <- read.csv("pml-testing.csv")
dim(traincsv)
dim(testcsv)
```
Here we have 160 vars and 19622 obs in the training set, and 20 for the test set.

## Cleaning the data
We must remove unnecessary variables starting with N/A's

```{r clean}
traincsv <- traincsv[,colMeans(is.na(traincsv))< .9] #remove the "mostly" na columns
traincsv <- traincsv[,-c(1:7)] #removing irrelavent metadata
```
removing near zero variance vars
```{r clean2}
nvz <- nearZeroVar(traincsv)
traincsv <- traincsv[,-nvz]
dim(traincsv)
```
Now we've finished removing unnecessary vars we can split the training set into validation adn sub training, the testing will be left alone used for final quiz test cases
```{r splitdata}
inTrain <- createDataPartition(traincsv$classe, p=0.7, list=F)
train <- traincsv[inTrain,]
valid <- traincsv[-inTrain,]
```
## Creating and testing models
Here's the bread and butterm we will test a few models, decision trees, random forest, gradiant boosted trees, and svm. This is probably more than we'd need to test but for good practice and comparison we will run them.

Set up control for training to use 3-fold cross validation
```{r control}
control <- trainControl(method="cv", number=3, verboseIter=F)
```

## Decision tree
**model:
```{r decision tree model}
mod_trees <- train(classe~., data=train, method="rpart", trControl = control, tuneLength = 5)
fancyRpartPlot(mod_trees$finalModel)
```
**Prediction
```{r decision tree prediction}
pred_trees <- predict(mod_trees, valid)
cmtrees <- confusionMatrix(pred_trees, factor(valid$classe))
cmtrees
```
## Random Forest
```{r random forest}
mod_rf <- train(classe~., data=train, method="rf", trControl = control, tuneLength = 5)

pred_rf <- predict(mod_rf, valid)
cmrf <- confusionMatrix(pred_rf, factor(valid$classe))
cmrf
```
##Gradient Boosted Trees
```{r Gbm}
mod_gbm <- train(classe~., data=train, method="gbm", trControl= control, tuneLength = 5, verbose = F)

pred_gbm <- predict(mod_gbm, valid)
cmgbm <- confusionMatrix(pred_gbm, factor(valid$classe))
cmgbm
```
##Support Vector Machine
```{r svm}
mod_svm <- train(classe~., data=train, method="svmLinear", trControl = control, tuneLength = 5, verbose = F)

pred_svm <- predict(mod_svm, valid)
cmsvm <- confusionMatrix(pred_svm, factor(valid$classe))
cmsvm
```
## Results
```{r resulsmatrix}
results<-matrix(c(0.537, 0.996, 0.992, 0.781, 0.463, 0.004, 0.008, 0.219), ncol = 2, nrow = 4)
colnames(results) <- c('Accuracy', 'OOS error')
rownames(results) <- c("Tree", 'RF', 'GBM', 'SVM')
print(results)
```

**The best fitting model for this set is the Random Forest model with 99% accuracy and .004 sample error, more than sufficent for testing.
```{r testing}
# predict on test set
preds <- predict(mod_rf, testcsv)
print(preds)
```
##Appendix
correlation matrix of variables in the training set
```{r corrmatrix}
correlationPlot <- cor(train[, -length(names(train))])
corrplot(correlationPlot, method = "color")
```
**Plotting the models
Twrees
```{r plot_trees, echo=FALSE}
plot(mod_trees)
```
Random forest
```{r plot_rf, echo=FALSE}
plot(mod_rf)
```
Gradient Boosted Trees
```{r plot_gbm, echo=FALSE}
plot(mod_gbm)
```

